\section{Results}

All models were trained using all four row groups from \texttt{"smartbkg\_dataset\_4k\_training.parquet"}, with a 75\%-25\% train-validation split. 
Evaluation was performed on all four row groups from \texttt{"smartbkg\_dataset\_4k\_testing.parquet"}.
Each model was trained for 10 epochs with a batch size of 256 and 32 hidden units. 

\begin{itemize}
    \item deepset \ref{model:DeepSet}
    \item deepset combined \ref{model:CombinedModel}
    \item deepset gcn \ref{model:DeepSet_wGCN}
    \item deepset combined with gcn \ref{model:CombinedModel_wGCN}
    \item transformer \ref{model:Transformer}
    \item deepset combined wwith gcn normalized \ref{model:CombinedModel_wGCN_Normalized}
\end{itemize}

\subsection{Model Evaluation}

To evaluate the performance of each model, we use the Receiver Operating 
Characteristic (ROC) curve, a widely adopted metric for binary 
classification tasks. The ROC curve plots the \textbf{True Positive Rate (TPR)} 
against the \textbf{False Positive Rate (FPR)}.

\paragraph{Definitions.}
\begin{itemize}
    \item \textbf{True Positive Rate (TPR):} It measures the proportion of actual positives that are correctly identified: 
    \[
    \text{TPR} = \frac{\text{TP}}{\text{TP} + \text{FN}}
    \]
    \item \textbf{False Positive Rate (FPR):} It measures the proportion of actual negatives that are incorrectly identified as positives:
    \[
    \text{FPR} = \frac{\text{FP}}{\text{FP} + \text{TN}}
    \]
\end{itemize}


The ideal model achieves a high TPR while maintaining a low FPR. 
The performance of a model can be quantified using the 
\textbf{Area Under the Curve (AUC)}. 
A higher AUC indicates better separability between the two classes.

\paragraph{Interpretation.}
\begin{itemize}
    \item AUC = 1.0: Perfect classifier.
    \item AUC = 0.5: Random guessing.
    \item AUC < 0.5: Model performs worse than random (potentially inverted labels).
\end{itemize}

\subsection{Optimal Model}
 

